---
title: "Data Wrangling in R, Lesson 3"
author: "Christy Garcia, Ph.D. and Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output:
  github_document: default
  html_notebook: default
---

## Introduction 
In the first two "Data Wrangling" sessions, we explored how to clean data, work with existing variables, and mutate variables in order to create new ones. This week we'll be taking a look at how to manage observations within a dataset. 

## Dependencies
This week we'll continue to use the super powerful `dplyr` package within `tidyverse`. We also need the `readr` and `here` packages again:

```{r load-packages}
# tidyverse packages
library(dplyr)     # data wrangling
library(readr)     # read and write csv files

# manage file paths
library(here)      # manage file paths
```

## Read in Data
This week we'll use the same two datasets as the last session -- the clean versions of `mpg` and `starwars`. Our first step is to read those in: 

```{r read}
mpg <- read_csv(here("data", "mpg.csv"))
starwars <- read_csv(here("data", "starwars.csv"))
```

## Review - Subsetting columns with `select`

Depending on where you get your data from, you might end up with a dataset that has hundreds or even thousands of variables. Chances are you are not actually going to use all of these variables. The `select()` function allows you to select which variables you would like to include: 

```{r select-mpg-1}
mpg_select <- select(mpg, manufacturer, model, year, cty, hwy)
```

If the variables/columns you want to select happen to be right next to each other, you can also tell it the range of columns to include: 

```{r select-mpg-2}
mpg_select2 <- select(mpg, manufacturer:year)
```

If you'd like to preserve all of the variables, but re-oder so that certain columns appear at the beginning of the data frame, you can embed the `everything()` function within `select()`. 

```{r mpg-select-3}
mpg_select3 <- select(mpg, cty, hwy, everything())
```

Try using `select()` with the `starwars` data frame. (For example, selecting only those variables that are related to phenotype.) 

```{r}
starwars_select <- select(starwars, name, hair_color, skin_color, eye_color)
```

## Subsetting rows with `filter`
The function `filter()` within the `dplyr` package allows us to subset observations based on a given variable value. We need to specify the following within the function: 

1. the name of the data frame
2. which variables we would like to filter by
3. what values of those variables we would like to select

For example, say we want to filter out the automobiles from the `mpg` data frame that are Audis and have four cylinders: 

```{r filter-mpg-1}
filter(mpg, manufacturer == "audi", cyl == 4)
```

(Note the quotes for the character value "audi" and the lack of quotes for the integer value for cylinders.)

If we do nothing else, this only prints the filtered results, but we can also save them to a new data frame if we want to later do some analysis of only this subset of the data. 

```{r filter-mpg-2}
audi4 <- filter(mpg, manufacturer == "audi", cyl == 4)
```

Practice using the `filter()` function by making a subset of the `starwars` data. (For example, filtering only the humans from Tatooine.)

```{r filter-starwars}
humanTat <- filter(starwars, homeworld == "Tatooine", species == "Human")
```

Like we saw last time, we can use operators to get a little bit more complicated. Here are the basic operators we have at our disposal: 

`>` greater than
`>=` greater than or equal to 
`<` less than
`<=` less than or equal to 
`!=` not equal to 
`==` equal to 
`&` and
`|` or
`!` not 

For example, let's `filter()` all automobiles that have four or six cylinders: 

```{r filter-mpg-3}
filter(mpg, cyl==4 | cyl ==6)
```

As we saw last time, we can also use the `%in%` operator to select multiple values to filter:

```{r filter-mpg-4}
filter(mpg, cyl %in% c(4, 6))
```

In this example, we are filtering all automobiles that are more than 6 cylinders and made before 2008: 

```{r filter-mpg-5}
filter(mpg, cyl >= 6 & year < 2008)
```

Yout turn! Try using `filter()` with some operators to filter the `starwars` data frame in two or three different ways. (For example, filtering the tall males, or those who have blue or blue-gray eyes and are from anywhere but Tatooine.) 

```{r filter-starwars-2}
filter(starwars, height >= 180 & gender == "male")
filter(starwars, homeworld != "Tatooine" & eye_color %in% c("blue", "blue-gray"))
```


## Choose rows by position with `slice`

To instead subset your data by row position, you can use the `slice()` function to include only certain observations. All you have to specify is the data frame to be sliced and which rows you would like to include: 

```{r silce-mpg}
mpg_sliced <- slice(mpg, 100:120)
```

Remember you can also use `head()` and `tail()` to get a quick look at the first six and last six observations, respectively. 

```{r head-mpg}
head(mpg)
```

```{r tail-mpg}
tail(mpg)
```

Slice and dice in the `starwars` data frame: 

```{r silce-starwars}
starwars_sliced <- slice(starwars, 1:20)
```

## Arrange rows with `arrange`

The function `arrange()` changes the order of rows within a data frame. You can think of it as analogous to the 'sort' function in Excel. First, we tell it what data frame we would like to re-order and then which variables(s) to order by: 

```{r arrange-mpg-1}
arrange(mpg, cty)
```

Just like with `filter()` and `slice()`, this new sorted order is not saved to the data frame unless you use the assignment operator: 

```{r arrange-mpg-2}
mpg_sorted <- arrange(mpg, cty)
```

Also, the default is ascending order unless you specify that you want descending order with `desc()`:

```{r arrange-mpg-3}
arrange(mpg, desc(cty))
```

*Note: Missing values (which we don't have any of!) are sorted to the end. 

If you specify more than one variable within the function, `arrange()` will sort by that variable after the preceding one (similar to 'sort/then by' in Excel).

```{r arrange-mpg-4}
arrange(mpg, cty, class)
```

Try re-ordering the rows of the `starwars` data frame. (For example, re-order by homeworld and then alphabetically by first name.) 

```{r arrange-starwars}
arrange(starwars, homeworld, name)
```

## Random Samples 
Sometimes it is useful to create a random sample of our data. This can use particularly helpful if you want to work out the kinks of a data cleaning or analysis process on a subset of the data, or if you want to create a subsample for additional analyses. 

```{r random-mpg}
# create sample
mpg_random <- sample_n(mpg, size = 50)
```

If you wanted every draw to have the same observations, you can set the "seed" (using `base::set.seed()`). In this case, it ensures that we all end up with the same random draw from the `mpg` data:

```{r random-seed-mpg}
# set seed
set.seed(12345)

# create sample
mpg_random2 <- sample_n(mpg, size = 50)
```

Try taking a random sample of 40 observations from the `starwars` data frame. There is no need to `set.seed()` here:

```{r sample-starwars}
starwars_sample <- sample_n(starwars, size = 40)
```
